Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2020-08-26 manuscript/plos/review_1.tex
updated figures and added supplementary tables

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study six network methods which identify subnetworks with high association scores to a phenotype. Specifically, we examine their utility to discover new biomarkers for breast cancer susceptibility by interrogating a genome-wide association study (GWAS) focused on French women with a family history of breast cancer and tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We perform an in-depth benchmarking of the methods with regards to the size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. For instance, we show a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Importantly, we find a significantly large overlap between the genes in the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Yet, network methods are notably unstable, producing different results when the input data changes slightly. To account for that, we produce a stable consensus subnetwork, formed by the most consistently selected genes. The stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than average. The network seems organized around \emph{CUL3}, encoding a ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely a lack of statistical power and interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

@@

We applied six network methods to the GENESIS dataset (Section~\ref{methods:methods}). As none of the networks examined by LEAN was significant (Benjamini-Hochberg [BH] correction adjusted P\=/value < 0.05), we obtained six solutions (Fig~\ref{fig:solution_overview}): one for each of the remaining four gene-based methods, one for SConES GI (which works at the SNP level), and the consensus. These solutions differ in many aspects, making it hard to draw joint conclusions. For starters, the overlap between the genes featured in each solution is quite small (Fig~\ref{fig:solution_overview}B). Another prominent difference is their size: the largest solution, produced by HotNet2, contains 440 genes, while heinz's contained only 4 genes. While SConES GI failed to recover any protein coding gene, by dealing with SNP networks it retrieved four subnetworks in intergenic regions, and another one overlapping an RNA gene (\emph{RNU6-420P}). Their topologies also differ, as measured by the median centrality (Table~\ref{tab:gene_solutions}) and the number of connected components (Fig~\ref{fig:solution_overview}D). Only two methods have more than one connected component: SConES, as described above, and HotNet2. HotNet2 produced 135 subnetworks, 115 of which have less than five genes. The second largest subnetwork (13 nodes) contains the two breast cancer susceptibility genes \emph{CASP8} and \emph{BLM}. Lastly, a pathway enrichment analysis (Section \ref{methods:pathway_enrichment}) also showed similarities and differences in the underlying mechanisms. It linked different parts of SigMod's solution network to four processes: protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~<~0.03). Interestingly, the dmGWAS solution is also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). But, additionally, it includes submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~<~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution are involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~<~0.04).

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[height=\textwidth]{./figures/figure_1.png}
  \caption{{\bf Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. } As LEAN did not produce any significant gene (BH adjusted P\=/value~<~0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)} Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54~\texttimes{}~10\textsuperscript{-7} for SNPs, 1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{(B)} Overlap between the genes selected by each of the methods, measured by Pearson correlation between indicator vectors. \textbf{(C)} Precision and recall of the evaluated methods with respect to Bonferroni-significant SNPs/genes in BCAC. For reference, we added a gray line with a slope of 1. \textbf{(D)} Solution networks produced by the different methods.}
  \label{fig:solution_overview}
\end{figure}

@@

A pathway enrichment analysis of the genes in the consensus network also shows similar pathways as the individual solutions. We found two involved mechanisms: \emph{mitochondrial translation} and \emph{attenuation phase}. The former is supported by genes like \emph{MRPS30} (VEGAS2 P\=/value~=~0.001), which encode a mitochondrial ribosomal protein and was also linked to breast cancer susceptibility \cite{quigley_5p12_2014}. Interestingly, increased mitochondrial translation has been found in cancer cells \cite{Yu2016Repositioning}, and its inhibition proposed as a therapeutic target. With regards to the attenuation phase of heat shock response, it involves three Hsp70 chaperones: HSPA1A, HSPA1B, and HSPA1L. The genes encoding these proteins are all near each other at 6p21, in the region known as HLA. In fact, out of the 22 SNPs that map to any of these three genes, 9 map to all of them, and 4 to two, making it hard to disentangle their effects. \emph{HSPA1A} was the most strongly associated gene (VEGAS2 P\=/value~=~8.37~\texttimes{}~10\textsuperscript{-4}).

@@

As shown above, and despite their similarities, the network methods produced remarkably different solutions. This is due to the particularities of each method, and directly or indirectly provide information about the dataset. For instance, the fact that LEAN did not return any biomarker implies that there is no gene such that both itself and its environment are on average strongly associated with the disease.

@@

DmGWAS carries out this process on every gene in the network. We used the implementation of dmGWAS in the dmGWAS 3.0 R package \cite{dmgwas}. We used the suggested hyperparameters $d = 2$ and $r = 0.1$. We used the function \emph{simpleChoose} to select the solution subnetwork, which aggregates the top 1\% subnetworks.
\end{description}

@@

We searched for pathways enriched in the gene subnetworks produced by the above methods. We conducted a hypergeometric test on pathways from Reactome \cite{Jassal2019} using the R package ReactomePA \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section \ref{methods:node_score}). We adjusted the P-values for multiple testing as in Benjamini and Hochberg \cite{Benjamini1995} (BH). Pathways with a BH adjusted P\=/value \textless~0.05 were deemed significant. As the significant pathways are often overlapping and redundant, the results were manually curated afterward.

@@

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study six network methods which identify subnetworks with high association scores to a phenotype. Specifically, we examine their utility to discover new biomarkers for breast cancer susceptibility by interrogating a genome-wide association study (GWAS) focused on French women with a family history of breast cancer and tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We perform an in-depth benchmarking of the methods with regards to the size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. For instance, we show a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Importantly, we find a significantly large overlap between the genes in the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Yet, network methods are notably unstable: the average Pearson correlation across all the methods between the genes selected on different subsamples of the data is 0.45, versus 0.57 simply using Bonferroni correction. To account for that, we produce a stable consensus subnetwork, formed by the most consistently selected genes. The stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than average. The network seems organized around \emph{CUL3}, encoding a ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely a lack of statistical power and interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

@@

We applied six network methods to the GENESIS dataset (Section~\ref{methods:methods}). As none of the networks examined by LEAN was significant (Benjamini-Hochberg [BH] correction adjusted P\=/value < 0.05), we obtained six solutions (Fig~\ref{fig:solution_overview}): one for each of the remaining four gene-based methods, one for SConES GI (which works at the SNP level), and the consensus. These solutions differ in many aspects, making it hard to draw joint conclusions. For starters, the overlap between the genes featured in each solution is quite small (Fig~\ref{fig:solution_overview}B). Another prominent difference is their size: the largest solution, produced by HotNet2, contains 440 genes, while heinz's contained only 4 genes. While SConES GI failed to recover any protein coding gene, by dealing with SNP networks it retrieved four subnetworks in intergenic regions, and another one overlapping an RNA gene (\emph{RNU6-420P}). Their topologies also differ, as measured by the median centrality (Table~\ref{tab:gene_solutions}) and the number of connected components (Fig~\ref{fig:solution_overview}D). Only two methods have more than one connected component: SConES, as described above, and HotNet2. HotNet2 produced 135 subnetworks, 115 of which have less than five genes. The second largest subnetwork (13 nodes) contains the two breast cancer susceptibility genes \emph{CASP8} and \emph{BLM}. Lastly, a pathway enrichment analysis (Section \ref{methods:pathway_enrichment}) also showed similarities and differences in the underlying mechanisms. It linked different parts of SigMod's solution network to four processes (\nameref{stab:sigmod_pwy}): protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~<~0.03). Interestingly, the dmGWAS solution (\nameref{stab:dmgwas_pwy}) is also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). But, additionally, it includes submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~<~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution (\nameref{stab:hotnet2_pwy}) are involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~<~0.04).

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[height=\textwidth]{./figures/figure_1.png}
  \caption{{\bf Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. } As LEAN did not produce any significant network (BH adjusted P\=/value~<~0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)} Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54~\texttimes{}~10\textsuperscript{-7} for SNPs, 1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{(B)} Overlap between the genes selected by each of the methods, measured by Pearson correlation between indicator vectors. \textbf{(C)} Precision and recall of the evaluated methods with respect to Bonferroni-significant SNPs/genes in BCAC. For reference, we added a gray line with a slope of 1. \textbf{(D)} Solution networks produced by the different methods.}
  \label{fig:solution_overview}
\end{figure}

@@

A pathway enrichment analysis of the genes in the consensus network also shows similar pathways as the individual solutions (\nameref{stab:consensus_pwy}). We found two involved mechanisms: \emph{mitochondrial translation} and \emph{attenuation phase}. The former is supported by genes like \emph{MRPS30} (VEGAS2 P\=/value~=~0.001), which encode a mitochondrial ribosomal protein and was also linked to breast cancer susceptibility \cite{quigley_5p12_2014}. Interestingly, increased mitochondrial translation has been found in cancer cells \cite{Yu2016Repositioning}, and its inhibition proposed as a therapeutic target. With regards to the attenuation phase of heat shock response, it involves three Hsp70 chaperones: HSPA1A, HSPA1B, and HSPA1L. The genes encoding these proteins are all near each other at 6p21, in the region known as HLA. In fact, out of the 22 SNPs that map to any of these three genes, 9 map to all of them, and 4 to two, making it hard to disentangle their effects. \emph{HSPA1A} was the most strongly associated gene (VEGAS2 P\=/value~=~8.37~\texttimes{}~10\textsuperscript{-4}).

@@

As shown above, and despite their similarities, the network methods produced remarkably different solutions. This is due to the particularities of each method, and directly or indirectly provide information about the dataset. For instance, the fact that LEAN did not return any gene implies that there is no gene such that both itself and its environment are on average strongly associated with the disease.

@@

DmGWAS carries out this process on every gene in the network. We used the implementation of dmGWAS in the dmGWAS 3.0 R package \cite{dmgwas}. We used the suggested hyperparameters $d = 2$ and $r = 0.1$. We used the function \texttt{simpleChoose} to select the solution subnetwork, which aggregates the top 1\% subnetworks.
\end{description}

@@

We searched for pathways enriched in the gene subnetworks produced by the above methods. We conducted a hypergeometric test on pathways from Reactome \cite{Jassal2019} using the function \texttt{enrichPathway} from the ReactomePA R package \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section \ref{methods:node_score}). We adjusted the P-values for multiple testing as in Benjamini and Hochberg \cite{Benjamini1995} (BH). Pathways with a BH adjusted P\=/value \textless~0.05 were deemed significant. As the significant pathways are often overlapping and redundant, the results were manually curated afterward.

@@

\paragraph*{S3 Table.}
\label{stab:sigmod_pwy}
{\bf Pathway enrichment analyses of the genes in SigMod solution.}

\paragraph*{S4 Table.}
\label{stab:dmgwas_pwy}
{\bf Pathway enrichment analyses of the genes in dmGWAS solution.}

\paragraph*{S5 Table.}
\label{stab:hotnet2_pwy}
{\bf Pathway enrichment analyses of the genes in HotNet2 solution.}

\paragraph*{S6 Table.}
\label{stab:consensus_pwy}
{\bf Pathway enrichment analyses of the genes in the consensus network.}
